---
title: "CCR7 analysis"
author: "Health Data Unit"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---


# Intro

This report contain documentation of the analysis of CD8 T cells from chronic HCV
patients.


```{r setup, include=FALSE}
## Dependencies
library(ggplot2)
library(tidyverse)
library(dplyr)
library(Nebulosa)
library(rmarkdown)
library(Seurat)

set.seed(333)

knitr::opts_chunk$set(
        warning = FALSE, 
        message = FALSE,
        cache = FALSE,
        fig.align='center'
)

## Setting paths
path2project <- '/media/ag-cherrmann/cramirez/tcd8ExscSeq/'
knitr::opts_knit$set(
        root.dir = path2project
)

## Interesting genes
genes.sel <- c('IL7R', 'TCF7', 'EOMES', 'TBX21', 'SELL', 'BCL2', 'BCL6',
               'STAT3', 'STAT4', 'PRDM1')
genes.sel <- toupper(genes.sel) 
```


## Visualization of TFs



 * Checking if genes are present in the count matrix.

```{r, checking_count_matrix}
## Maike Hoffman data preprocessing
## gene names must be standardized
himmer <- read.csv('data/maike2020/nina_thimme_raw_counts.csv', 
                   header = TRUE)
r_names <- himmer$X
himmer <- dplyr::select(himmer, -X)
himmer_mtx <- apply(himmer, 2, as.numeric)
rownames(himmer_mtx) <- gsub('__chr.*', '', r_names) 

#sapply(genes.sel, function(x) x %in% toupper(rownames(himmer_mtx)))
```

Genes are present in the initial matrix but lost in a further step.

 * Then, normalized gene expression of the markers in the initial uncorrected matrix were projected
 onto the corrected UMAPs as shown in the next plots.
 
 
# Joint expression of markers with CCR7

## Densities 

```{r, densities, fig.width=14, fig.height=14}
seurat <- readRDS('data/maike2020/hofmann_hcv_seu.rds')
seurat <- subset(seurat, nina_annotations %in% c('mem', 'exh', 'trans'))
#seurat <- UpdateSeuratObject(seurat)
cell_types <- DimPlot(seurat, 
                      group.by = 'nina_annotations', 
                      pt.size = 2.5, 
                      combine = FALSE)


## Creating an additional seurat object to store RNA
## uncorrected data
seurat.rna <- CreateSeuratObject(
  counts = himmer_mtx,
  project = 'HBV_RNA', 
  assay = 'RNA', 
  min.cells = 0, 
  min.features = 0,
  meta.data = seurat@meta.data
)

## Checking cells are the same and in the same order
#any(colnames(seurat) != colnames(seurat.rna))
#[1] FALSE       ## They are!

## Normalization of the non-integrated data
seurat.rna <- ScaleData(seurat.rna) %>% NormalizeData()
## Extraction of the markers
sapply(genes.sel, function(x) x %in% rownames(seurat.rna))
markers.rna <- FetchData(seurat.rna, vars = genes.sel) 
colnames(markers.rna) <- paste0('rna_', colnames(markers.rna))

## Adding markers to the corrected UMAP representation
#seurat@meta.data <- cbind(seurat@meta.data, markers.rna)
for (gene in genes.sel){
  FeaturePlot(seurat, 
             features = c('rna_', paste0('rna_', gene)),
             pt.size = 2.5, 
             combine = FALSE)
}

names(genes.sel) <- genes.sel
density_plots <- lapply(genes.sel, function(gene){
  plot_density(seurat, 
             features = c('rna_CCR7', paste0('rna_', gene)),
             size = 2.5, 
             joint = TRUE, 
             combine = FALSE)[[3]] +
    ggtitle(paste0('CCR7+', '', gene, '+')) +
    theme_void()
})

markers.list <- c(cell_types, density_plots)
gridExtra::grid.arrange(
        grobs=markers.list,
        ncol=3
)
```


## Gene expression

```{r, gene_expression, fig.width=14, fig.height=20}
plots <- list()
for (gene in genes.sel) {
  new.plot <- FeaturePlot(seurat, 
            features = c('rna_CCR7', paste0('rna_', gene)),
            pt.size = 2.5, 
            blend = TRUE,
            blend.threshold = 0.05, 
            combine = FALSE)
  plots <- c(plots, new.plot)
}
length(plots)


gridExtra::grid.arrange(
  grobs = plots,
  ncol=4
)
```
