---
title: "Gene regulation in CD8 T cells from chronic HCV patients"
author: "Health Data Unit"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---


# Intro

This report contain documentation of the analysis of CD8 T cells from chronic HCV
patients.


```{r setup, include=FALSE}
## Dependencies
library(ggplot2)
library(tidyverse)
library(dplyr)
library(Nebulosa)
library(rmarkdown)
library(igraph)
library(rio)
library(Seurat)
library(readxl)
library(dittoSeq)
library(viridis)
library(reshape2)
library(pander)

set.seed(333)

knitr::opts_chunk$set(
        warning = FALSE, 
        message = FALSE,
        cache = FALSE,
        fig.align='center'
)

## Setting paths
path2project <- '/media/ag-cherrmann/cramirez/tcd8ExscSeq/'
knitr::opts_knit$set(
        root.dir = path2project
)

## Interesting genes
genes.sel <- c('EZH2', 'Id3', 'Id2', 'KDM6A', 'PRDM1', 'Tcf7')
genes.sel <- toupper(genes.sel) 
```


## Visualization of TFs



 * Checking if genes are present in the count matrix.

```{r, checking_count_matrix}
## Maike Hoffman data preprocessing
## gene names must be standardized
himmer <- read.csv('data/maike2020/nina_thimme_raw_counts.csv', 
                   header = TRUE)
r_names <- himmer$X
himmer <- dplyr::select(himmer, -X)
himmer_mtx <- apply(himmer, 2, as.numeric)
rownames(himmer_mtx) <- gsub('__chr.*', '', r_names) 

sapply(genes.sel, function(x) x %in% toupper(rownames(himmer_mtx)))
```

Genes are present in the initial matrix but lost in a further step.

 * Then, normalized gene expression of the markers in the initial uncorrected matrix were projected
 onto the corrected UMAPs as shown in the next plots.
 
 
 ### Normalized expression

```{r, fig.width=12, fig.height=11}
seurat <- readRDS('data/maike2020/hofmann_hcv_seu.rds')
seurat <- subset(seurat, nina_annotations %in% c('mem', 'exh'))
#seurat <- UpdateSeuratObject(seurat)
cell_types <- DimPlot(seurat, 
                      group.by = 'nina_annotations', 
                      pt.size = 2.5, 
                      combine = FALSE)


## Creating an additional seurat object to store RNA
## uncorrected data
seurat.rna <- CreateSeuratObject(
  counts = seurat@assays$RNA@counts,
  project = 'HBV_RNA', 
  assay = 'RNA', 
  min.cells = 0, 
  min.features = 0,
  meta.data = seurat@meta.data
)

## Checking cells are the same and in the same order
#any(colnames(seurat) != colnames(seurat.rna))
#[1] FALSE       ## They are!

## Normalization of the non-integrated data
seurat.rna <- ScaleData(seurat.rna) %>% NormalizeData()
## Extraction of the markers
markers.rna <- FetchData(seurat.rna, vars = genes.sel) 
colnames(markers.rna) <- paste0('rna_', colnames(markers.rna))

## Adding markers to the corrected UMAP representation
seurat@meta.data <- cbind(seurat@meta.data, markers.rna)
markers <- FeaturePlot(seurat, 
             features = paste0('rna_', genes.sel),
             pt.size = 2.5, 
             combine = FALSE)
for (i in 1:length(markers)) {
        markers[i][[1]] <- markers[i][[1]] +
                                theme_void() +
                              scale_color_viridis()
}
markers.list <- c(cell_types, markers)

gridExtra::grid.arrange(
        grobs=markers.list,
        ncol=3
)
```


### UTX gene expression


```{r, utx_violin_plot, fig.align='center', fig.width=10}
utx.vln <- dittoPlot(subset(seurat, rna_KDM6A > 0), "rna_KDM6A", 
          group.by = "nina_annotations",
    plots = c("vlnplot", "jitter", "boxplot"),
    jitter.color = "blue", jitter.size = 0.5,
    boxplot.color = "white", boxplot.width = 0.1,
    boxplot.fill = FALSE,
    vlnplot.scaling = "count"
    ) + 
  labs(x='', y='UTX Normalized gene expression') +
  theme(legend.position = 'none',
        axis.text.x = element_text(size = 16)) +
  scale_fill_manual(values=c('salmon', 'cyan3'))

utx.dotplot <- DotPlot(seurat, 
                       features = 'rna_KDM6A',
                       group.by = 'nina_annotations',
                       dot.scale = 15) +
              scale_color_viridis() +
              labs(x='', y='')

gridExtra::grid.arrange(
  grobs=list(utx.vln, utx.dotplot), ncol=2
)

```



## GRN inferred using GRANET

GRANET is TFA inference framework based in the following steps: i) Inference of 
co-expressed genes to TFs using GRNBoost2 which defines modules, ii) Pruning the
modules using ATACSeq data.

The next graph shows the interactions containing the genes EZH2, ID3, ID2, KDM6A,
PRDM1, TCF7 (of which only ID2 and PRDM1) were found to be in the network
inferred using GRANET.



```{r}
network <- read.table('analysis/scenic_atac_specific_filtered_network.tsv',
                  header = TRUE)
#head(network)

network <- filter(network, TF %in% genes.sel | target %in% genes.sel)

## Definition of the nodes list
edges <- dplyr::select(network, TF, target)

## Definition of the edges list
nodes <- data.frame(node=unique(c(network$TF, network$target)))
#head(nodes)

#net.network <- network(edges, 
#                       vertex.attr = nodes)

net.igraph <- graph_from_data_frame(
        d = edges,
        vertices = nodes, 
        directed = FALSE
)

plot(net.igraph, 
     vertex.size=0.02, 
     layout=layout_nicely)
```


```{r, eval=FALSE}
tfa <- read.csv('analysis/scenic/hgv_3000/aucell.csv')
colnames(tfa) %>% sort
colnames(tfa) <- gsub('\\.', '', colnames(tfa))

## function to extract TFs in column 7 from SCENIC results
extract_targets <- function(x){
        targets <- regmatches(x, gregexpr("'[^']*'", x))[[1]] 
        gsub("'", '', targets)
}

#'LAG3' %in% rownames(seurat)
#sapply(genes.sel, function(x) x %in% rownames(seurat))



regulons <- read.csv('analysis/scenic/hgv_3000//reg.csv', skip = 2)
#head(regulons)

regulons.list <- lapply(1:nrow(regulons), 
       function(i){
               extract_targets(regulons$X.7[i])
})
names(regulons.list) <- regulons$TF

regulons.df.list <- lapply(1:length(regulons.list), 
                      function(i) {
                              df <- data.frame(
                                      target=regulons.list[i][[1]]
                              )
                              df$'tf' <- names(regulons.list)[i]
                              df[, c('tf', 'target')]
                      }) 

regulons.df <- do.call(rbind, regulons.df.list)

network <- filter(regulons.df, target %in% genes.sel)
network <- unique(network)
network %>%
        paged_table()
```

## Scoring the CD8 T cells from chronically infected HBV patients 


Then, we evaluated the presence of UTX related signatures as reported in [Mitchell J E et al, 2021](https://pubmed.ncbi.nlm.nih.gov/33852868/). The following UMAP projection show the CD8 T cell 
types and the densities for the upregulated, downregulated and all genes (union of upregulated and
downregulated) in the UTX signature.

NOTE: Gene names overlap between signatures and names in the seurat object were checked. See, the
following chunk for details.


## Signature scores (Densities)

```{r, fig.width=10}
## Scoring CD8 T cells with UTX signatures
if ( ! file.exists('data/1-s2.0-S2211124721002801-mmc2.xlsx')) {
  download.file('https://ars.els-cdn.com/content/image/1-s2.0-S2211124721002801-mmc2.xlsx',
              destfile = 'data/1-s2.0-S2211124721002801-mmc2.xlsx')
}
utx.sign.down <- read_xlsx('data/1-s2.0-S2211124721002801-mmc2.xlsx', 
                      sheet = 3, skip = 3) %>%
                    unlist() %>%
                       unname() %>% toupper()
utx.sign.up <- read_xlsx('data/1-s2.0-S2211124721002801-mmc2.xlsx', 
                      sheet = 4, skip = 3) %>%
                    unlist() %>%
                       unname() %>% toupper()
utx.sign <- read_xlsx('data/1-s2.0-S2211124721002801-mmc2.xlsx', 
                      sheet = 1, skip = 3)
utx.sign <- dplyr::select(utx.sign, Genes) %>%
                    unlist() %>%
                       unname() %>% toupper()
sign.list <- list(utx.sign.up=utx.sign.up, 
                  utx.sign.down=utx.sign.down,
                  utx.sign=utx.sign)

## Checking if mouse and human gene names match
#lapply(sign.list, function(x) length(intersect(x, rownames(seurat.rna))))
#$utx.sign.up
#[1] 202
#
#$utx.sign.down
#[1] 317
#
#$utx.sign
#[1] 12048

## Scoring cells with signatures
seurat.rna <- AddModuleScore(seurat.rna, 
                         features = sign.list, 
                         name = paste0(names(sign.list), '_'))
scores <- dplyr::select(seurat.rna@meta.data, `utx.sign.up_1`:`utx.sign_3`)
colnames(scores) <- c('UTX_sign_up', 'UTX_sign_down', 'UTX_signature_all')

## Adding scores to metatadata in the integrated UMAP
## checking cells are the same and in the same order
#any(colnames(seurat) != rownames(scores))
# [1] FALSE.  ## they are
seurat@meta.data <- cbind(seurat@meta.data, scores)

signatures <- plot_density(seurat, 
             features = colnames(scores),
             size = 2.5, 
             combine = FALSE)
for (i in 1:length(signatures)) {
        signatures[i][[1]] <- signatures[i][[1]] +
                                theme_void() 
}
signatures.list <- c(cell_types, signatures)

gridExtra::grid.arrange(
        grobs=signatures.list,
        ncol=2
)
```

## UTX scores


```{r, fig.width=10, fig.align='center'}
signatures <- FeaturePlot(seurat, 
             features = colnames(scores),
             pt.size = 2.5, 
             combine = FALSE)
for (i in 1:length(signatures)) {
        signatures[i][[1]] <- signatures[i][[1]] +
                                theme_void() + scale_color_viridis()
}
signatures.list <- c(cell_types, signatures)

gridExtra::grid.arrange(
        grobs=signatures.list,
        ncol=2
)
```


As can be observed, there is a mutually exclusive expression of the up and down-regulated gene
signatures. Interestingly, down-regulated genes in the UTX mutant have higher scores in the 
memory progenitor, suggesting that in the WT UTX is acting as an activator of many genes in 
this cluster.

From looking also UTX gene expression we can conclude that this factor could be expressed and
active in memory cells cluster.

```{r, violin_plot, fig.align='center'}
up.plot <- dittoPlot(seurat, "UTX_sign_up", 
          group.by = "nina_annotations",
    plots = c("vlnplot", "jitter", "boxplot"),
    jitter.color = "blue", jitter.size = 0.5,
    boxplot.color = "white", boxplot.width = 0.1,
    boxplot.fill = FALSE,
    vlnplot.scaling = "count"
    ) + 
  labs(x='', y='UTX Signature Score (Upregulated genes)') +
  ggtitle('Upregulated genes') +
  theme(legend.position = 'none',
        axis.text.x = element_text(size = 16)) +
  scale_fill_manual(values=c('salmon', 'cyan3'))

down.plot <- dittoPlot(seurat, "UTX_sign_down", 
          group.by = "nina_annotations",
    plots = c("vlnplot", "jitter", "boxplot"),
    jitter.color = "blue", jitter.size = 0.5,
    boxplot.color = "white", boxplot.width = 0.1,
    boxplot.fill = FALSE,
    vlnplot.scaling = "count"
    ) + 
  labs(x='', y='UTX Signature Score (Downregulated genes)') +
  ggtitle('Downregulated genes') +
  theme(legend.position = 'none',
        axis.text.x = element_text(size = 16)) +
  scale_fill_manual(values=c('salmon', 'cyan3'))

gridExtra::grid.arrange(
  grobs=list(up.plot, down.plot), ncol=2
)
```

From the violin plots it can be seen that the UTX signature score for upregulated genes is 
higher in memory clusters.

```{r}
DotPlot(seurat, 
        features = c("UTX_sign_up", "UTX_sign_down"),
        group.by = 'nina_annotations', dot.scale = 20) +
        scale_color_viridis(option = 'viridis') +
        labs(x='', y='')
```

