---
title: "Analysis of the regulons rewiring after treatment"
author: "Health Data Unit"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---


Analysis of the rewiring of regulons inferred based in scRNA-Seq conexpression and
chromatin open regions from bulk ATAC Seq.


```{r, setup, include=FALSE}
library(igraph)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(ggVennDiagram)
library(Seurat)
library(viridis)
```


## Loading regulons

The inference of the regulons is implemented in the reports/network_inference.Rmd file.


```{r, calculation_of_distances}
granet.interactions.df.list <- readRDS(
        '/media/ag-cherrmann/cramirez/tcd8ExscSeq/analysis/regulons/granet_interactions.rds'
)

## Intersection of Tfs across categories
reg.tfs <- lapply(granet.interactions.df.list, function(df) unique(df$TF))
shared.tfs <- unique(Reduce(intersect, reg.tfs)) %>% sort()
## Union of all genes 
reg.targets <- lapply(granet.interactions.df.list, function(df) unique(df$gene))
targets <- unique(unlist(reg.targets)) %>% sort()

## create a matrix of genes (Tfs) x genes (targets) per category
## 1 if there is a interaction otherwise 0
adj.mat <- matrix(0, 
                  nrow = length(shared.tfs), 
                  ncol = length(targets))
rownames(adj.mat) <- shared.tfs
colnames(adj.mat) <- targets
adj.mat.list <- list(adj.mat, adj.mat, adj.mat)
names(adj.mat.list) <- names(granet.interactions.df.list)
names(adj.mat.list)

## calculate the sum(abs(x, y)) where x and y are rows from the previous matrix
## in different categories
for ( cat in names(adj.mat.list) ){
        df <- filter(granet.interactions.df.list[cat][[1]], TF %in% shared.tfs)
        df <- filter(df, gene %in% targets)
        
        for (i in 1:nrow(df)){
                adj.mat.list[cat][[1]][df$TF[i], df$gene[i] ] <- 1
        }
}

dif.gainRev <- adj.mat.list$gained - adj.mat.list$reversed
ranking <- apply(dif.gainRev, 1, function(x) sum(abs(x))) %>% sort(decreasing = TRUE)
ranking.df <- data.frame(tf=names(ranking),
                         distance=ranking)
```



```{r, vis}
pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/rewiring_ranking.pdf',
    width = 5, height =5)
ranking.df %>%
        mutate(rank=1:nrow(ranking.df)) %>%
        mutate(gene_label=ifelse(rank<10 | tf %in% c('BACH2'), tf, '')) %>%
        ggplot(aes(x=rank, y=distance,
                   label=gene_label)) +
                geom_point(size=2,
                           colour='steelblue') +
                geom_text_repel(force = 3) +
                theme_classic() +
                labs(x='Rank', y='Regulon distance (Gained Vs Reversed)')
dev.off()
```


```{r, creating_igraph_object}
stat1.gained <- filter(granet.interactions.df.list$gained, TF=='STAT1')
stat1.gained$category <- 'gained'
stat1.reversed <- filter(granet.interactions.df.list$reversed, TF=='STAT1')
stat1.reversed$category <- 'reversed'
stat1.reg <- rbind(stat1.gained, stat1.reversed)

nodes.df <- data.frame(gene=unique(c('STAT1', stat1.reg$gene)))
edges.df <- data.frame(from=stat1.reg$TF, to=stat1.reg$gene, category=stat1.reg$category)
g <- graph_from_data_frame(edges.df, directed = TRUE, vertices = nodes.df)
```


# STAT1 example


```{r, fig.height=10, fig.width=10}
selected.genes <- c('FOXP1', 'IRS2', 'PRDM1',
                    'TCF7', 'TNFRSF19', 'GZMA',
                    'HAVCR2')
labels <- sapply(V(g)$name, function(x){ ifelse(x %in% selected.genes, x, '') })

colors <- plyr::mapvalues(
        E(g)$category,
        from = unique(E(g)$category),
        to = c('salmon', 'green')
) 

l = layout_nicely(g)

pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/stat1_rewiring.pdf')
plot(g,
     vertex.size=1.5,
     vertex.label.dist=1,
     vertex.label.color='black',
     vertex.label.cex=0.5,
     
     edge.arrow.size=0.5,
     edge.curved=.1,
     edge.color=colors,
     
     layout=l)
dev.off()
```

# Ratio of regulon sizes

```{r}
interactions.list <- lapply(names(granet.interactions.df.list), 
                            function(name)
                            mutate(granet.interactions.df.list[name][[1]], category=name))
interactions.df <- do.call(rbind, interactions.list)

freq.df <- interactions.df %>%
        select(TF, category) %>%
        table() 
top.50 <- freq.df %>% 
        apply(1, sum) %>%
        sort(decreasing = TRUE) %>%
        head(50) %>%
        names()

ratios <- freq.df  %>%
        as.matrix() %>%
        apply(1, function(x) x/sum(x)) %>%
        as.data.frame() %>%
        t() %>%
        as.data.frame()

order.regs <- arrange(ratios, desc(scarred)) %>%
                rownames()


pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/ratio_interactions_by_category.pdf',
    width = 5, height = 7)
freq.df %>%
        as.data.frame() %>%
        filter(TF %in% top.50) %>%
        mutate(TF=factor(TF, levels = order.regs)) %>%
                ggplot(aes(x=TF, y=Freq, fill=category)) +
                        geom_bar(position = 'fill', stat = 'identity') +
                        coord_flip() +
                        theme_bw() +
                        theme(panel.grid = element_blank()) +
                        labs(x='Regulon', y='Ratio of interactions',
                             fill='')  
dev.off()

pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/freq_interactions_by_category.pdf',
    width = 5, height = 7)
freq.df %>%
        as.data.frame() %>%
        filter(TF %in% top.50) %>%
        mutate(TF=factor(TF, levels = rev(top.50))) %>%
                ggplot(aes(x=TF, y=Freq, fill=category)) +
                        geom_bar(position = 'stack', stat = 'identity') +
                        coord_flip() +
                        theme_bw() +
                        theme(panel.grid = element_blank()) +
                        labs(x='Regulon', y='Number of interactions',
                             fill='')  
dev.off()
```


# Evaluation of the sign of the regulation


```{r}
interactions.df <- interactions.df %>%
        mutate(interaction=paste0(gene, '--',TF))
int.list <- split(interactions.df$interaction, interactions.df$category) 
ggVennDiagram(int.list)
```


```{r, reg_type_hub0044, fig.width=15, fig.height=24}
## calculate DEGs treated/untreated separately by patient
seurat$orig.ident %>% table()

## function to categorise genes (up-, down-regulated or non-significant)
add_regulation_type <- function(degs.dataframe){
        degs <- degs.dataframe
        degs$'regulation_type' <- ''
        
        for (i in 1:nrow(degs)) {
                #if ( 0.05 < degs$p_val[i] ) {
                #        degs$regulation_type[i] <- 'Non-significant'
                #} else 
                if ( degs$avg_log2FC[i] < 0 ) {
                          degs$regulation_type[i] <- 'Down'    
                      }  else if ( 0 < degs$avg_log2FC[i] ) {
                          degs$regulation_type[i] <- 'Up'    
                      } else if ( 0 == degs$avg_log2FC[i] ) {
                          degs$regulation_type[i] <- 'No change'    
                }
        }
        
        return(degs)
}


degs.hub0709 <- subset(seurat, 
                       orig.ident %in% c('HUBF0709BL', 'HUBF0709FUDay2'))  %>%
                        FindMarkers(ident.1 = 'cured',
                                    ident.2 = 'chronic',
                                    group.by = 'condition', 
                                    min.pct= 0, 
                                    logfc.threshold = 0)

degs.hub0044 <- subset(seurat, 
                       orig.ident %in% c('HUBF0044BL', 'HUBF0044SORT3BFU'))  %>%
                        FindMarkers(ident.1 = 'cured',
                                    ident.2 = 'chronic',
                                    group.by = 'condition', 
                                    min.pct= 0, 
                                    logfc.threshold = 0)

## Add regulation type
degs.hub0709 <- add_regulation_type(degs.hub0709)
degs.hub0044 <- add_regulation_type(degs.hub0044)




label_targets <- function(degs, int.df, peak_category){
        tfs <- unique(int.df$TF)
        labelled_targets <- list()
        
        for (i in 1:length(tfs)) {
                targets <- filter(int.df, 
                                  category == peak_category & 
                                          TF == tfs[i]) %>%
                                   pull(gene) 
                ## categorise the target (up-, down-regulated or non-significant)
                regulation_type <- degs[targets, ] %>% pull(regulation_type)
                names(regulation_type) <- targets
                labelled_targets[i][[1]] <- regulation_type
        }
        names(labelled_targets) <- tfs
        
        return(labelled_targets)
}

labelled_targets <- lapply( unique(interactions.df$category),
                            function(cat) {
                                   label_targets(degs = degs.hub0044, 
                                                 int.df = interactions.df, 
                                                 peak_category = cat) 
                            }
)
names(labelled_targets) <- unique(interactions.df$category)


labelled_targets.vector <- unlist(labelled_targets)
labelled_targets.df <- data.frame(regulation_type=labelled_targets.vector,
                                  category.tf.target=names(labelled_targets.vector))
labelled_targets.df$'category' <- sapply(labelled_targets.df$category.tf.target,
                                         function(st){strsplit(st, split = '\\.')[[1]][1] })
labelled_targets.df$'TF' <- sapply(labelled_targets.df$category.tf.target,
                                         function(st){strsplit(st, split = '\\.')[[1]][2] })
labelled_targets.df$'target' <- sapply(labelled_targets.df$category.tf.target,
                                         function(st){strsplit(st, split = '\\.')[[1]][3] })



peak_categories <- unique(labelled_targets.df$category)
names(peak_categories) <- peak_categories
freqs.regulation.type <- lapply(peak_categories,
       function(cat){
        freqs.df <- labelled_targets.df %>%
                filter(category==cat) %>%
                select(TF, regulation_type) %>%
                       table() %>%
                       as.data.frame()
        order.tf <- freqs.df %>%
                        filter(regulation_type=='Up') %>%
                                arrange(Freq) %>%
                                pull(TF) %>%
                                as.character()
        
        freqs.df %>% 
                mutate(TF=factor(TF, levels = unique(order.tf))) %>% 
                        ggplot(aes(x=TF, y=Freq,
                                fill=regulation_type)) +
                                geom_bar(stat = 'identity', position = 'stack') +
                                        scale_fill_manual(values = c('Down'='red',
                                                                     'No change'='gray30',
                                                                     'Up'='steelblue')) +
                        coord_flip() +
                        labs(x='', y='Frequency of targets', fill='',
                             title = cat) +
                        theme_classic() 
        
        
})



props.reg_type <- lapply(peak_categories,
       function(cat){
        freqs.df <- labelled_targets.df %>%
                filter(category==cat) %>%
                select(TF, regulation_type) %>%
                       table() %>%
                       as.data.frame()
        
        freqs.table <- labelled_targets.df %>%
                filter(category==cat) %>%
                select(TF, regulation_type) %>%
                       table()
        order.tf <- freqs.table %>%
                as.matrix() %>%
                apply(1, function(x) x/sum(x)) %>%
                t() %>%
                as.data.frame() %>%
                arrange(desc(Up)) %>%
                rownames()
        
        freqs.df %>% 
                mutate(TF=factor(TF, levels = unique(order.tf))) %>% 
                        ggplot(aes(x=TF, y=Freq,
                                fill=regulation_type)) +
                                geom_bar(stat = 'identity', position = 'fill') +
                                        scale_fill_manual(values = c('Down'='red',
                                                                     'No change'='gray30',
                                                                     'Up'='steelblue')) +
                        coord_flip() +
                        labs(x='', y='Frequency of targets', fill='',
                             title = cat) +
                        theme_classic() 
        
        
        
})

pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/regulons_target_regulation_hub0044.pdf',
    width = 15, height = 24)
gridExtra::grid.arrange(
        grobs = c(freqs.regulation.type,
                  props.reg_type), ncol = 3
)
dev.off()
```


## Regulatory types in regulons (hub0709)


```{r, reg_type_hub0709, fig.width=15, fig.height=24}

labelled_targets <- lapply( unique(interactions.df$category),
                            function(cat) {
                                   label_targets(degs = degs.hub0709, 
                                                 int.df = interactions.df, 
                                                 peak_category = cat) 
                            }
)
names(labelled_targets) <- unique(interactions.df$category)


labelled_targets.vector <- unlist(labelled_targets)
labelled_targets.df <- data.frame(regulation_type=labelled_targets.vector,
                                  category.tf.target=names(labelled_targets.vector))
labelled_targets.df$'category' <- sapply(labelled_targets.df$category.tf.target,
                                         function(st){strsplit(st, split = '\\.')[[1]][1] })
labelled_targets.df$'TF' <- sapply(labelled_targets.df$category.tf.target,
                                         function(st){strsplit(st, split = '\\.')[[1]][2] })
labelled_targets.df$'target' <- sapply(labelled_targets.df$category.tf.target,
                                         function(st){strsplit(st, split = '\\.')[[1]][3] })



peak_categories <- unique(labelled_targets.df$category)
names(peak_categories) <- peak_categories
freqs.regulation.type <- lapply(peak_categories,
       function(cat){
        freqs.df <- labelled_targets.df %>%
                filter(category==cat) %>%
                select(TF, regulation_type) %>%
                       table() %>%
                       as.data.frame()
        order.tf <- freqs.df %>%
                        filter(regulation_type=='Up') %>%
                                arrange(Freq) %>%
                                pull(TF) %>%
                                as.character()
        
        freqs.df %>% 
                mutate(TF=factor(TF, levels = unique(order.tf))) %>% 
                        ggplot(aes(x=TF, y=Freq,
                                fill=regulation_type)) +
                                geom_bar(stat = 'identity', position = 'stack') +
                                        scale_fill_manual(values = c('Down'='red',
                                                                     'No change'='gray30',
                                                                     'Up'='steelblue')) +
                        coord_flip() +
                        labs(x='', y='Frequency of targets', fill='',
                             title = cat) +
                        theme_classic() 
        
        
})

gridExtra::grid.arrange(
        grobs = freqs.regulation.type, ncol = 3
)


props.reg_type <- lapply(peak_categories,
       function(cat){
        freqs.df <- labelled_targets.df %>%
                filter(category==cat) %>%
                select(TF, regulation_type) %>%
                       table() %>%
                       as.data.frame()
        
        freqs.table <- labelled_targets.df %>%
                filter(category==cat) %>%
                select(TF, regulation_type) %>%
                       table()
        order.tf <- freqs.table %>%
                as.matrix() %>%
                apply(1, function(x) x/sum(x)) %>%
                t() %>%
                as.data.frame() %>%
                arrange(desc(Up)) %>%
                rownames()
        
        freqs.df %>% 
                mutate(TF=factor(TF, levels = unique(order.tf))) %>% 
                        ggplot(aes(x=TF, y=Freq,
                                fill=regulation_type)) +
                                geom_bar(stat = 'identity', position = 'fill') +
                                        scale_fill_manual(values = c('Down'='red',
                                                                     'No change'='gray30',
                                                                     'Up'='steelblue')) +
                        coord_flip() +
                        labs(x='', y='Frequency of targets', fill='',
                             title = cat) +
                        theme_classic() 
        
        
        
})

pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/regulons_target_regulation_hub0709.pdf',
    width = 15, height = 24)
gridExtra::grid.arrange(
        grobs = c(freqs.regulation.type,
                  props.reg_type), ncol = 3
)
dev.off()
```

# Calculcation of the fraction of positive or negative interactions

```{r}
lapply(peak_categories,
       function(cat){
        freqs.df <- labelled_targets.df %>%
                filter(category==cat) %>%
                select(TF, regulation_type) %>%
                       table() %>%
                        as.matrix() %>%
                        apply(2, sum)
        prop <- freqs.df / sum(freqs.df)
        prop
})
```


