---
title: "Analysis of the regulons rewiring after treatment"
author: "Health Data Unit"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
  pdf_document: default
---


Analysis of the rewiring of regulons inferred based in scRNA-Seq conexpression and
chromatin open regions from bulk ATAC Seq.


```{r, setup, include=FALSE}
library(igraph)
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(ggVennDiagram)
library(Seurat)
library(viridis)
```


## Loading regulons

The inference of the regulons is implemented in the reports/network_inference.Rmd file.


```{r, calculation_of_distances}
granet.interactions.df.list <- readRDS(
        '/media/ag-cherrmann/cramirez/tcd8ExscSeq/analysis/regulons/granet_interactions.rds'
)

## Intersection of Tfs across categories
reg.tfs <- lapply(granet.interactions.df.list, function(df) unique(df$TF))
shared.tfs <- unique(Reduce(intersect, reg.tfs)) %>% sort()
## Union of all genes 
reg.targets <- lapply(granet.interactions.df.list, function(df) unique(df$gene))
targets <- unique(unlist(reg.targets)) %>% sort()

## create a matrix of genes (Tfs) x genes (targets) per category
## 1 if there is a interaction otherwise 0
adj.mat <- matrix(0, 
                  nrow = length(shared.tfs), 
                  ncol = length(targets))
rownames(adj.mat) <- shared.tfs
colnames(adj.mat) <- targets
adj.mat.list <- list(adj.mat, adj.mat, adj.mat)
names(adj.mat.list) <- names(granet.interactions.df.list)
names(adj.mat.list)

## calculate the sum(abs(x, y)) where x and y are rows from the previous matrix
## in different categories
for ( cat in names(adj.mat.list) ){
        df <- filter(granet.interactions.df.list[cat][[1]], TF %in% shared.tfs)
        df <- filter(df, gene %in% targets)
        
        for (i in 1:nrow(df)){
                adj.mat.list[cat][[1]][df$TF[i], df$gene[i] ] <- 1
        }
}

dif.gainRev <- adj.mat.list$gained - adj.mat.list$reversed
ranking <- apply(dif.gainRev, 1, function(x) sum(abs(x))) %>% sort(decreasing = TRUE)
ranking.df <- data.frame(tf=names(ranking),
                         distance=ranking)
```



```{r, vis}
pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/rewiring_ranking.pdf',
    width = 5, height =5)
ranking.df %>%
        mutate(rank=1:nrow(ranking.df)) %>%
        mutate(gene_label=ifelse(rank<10 | tf %in% c('BACH2'), tf, '')) %>%
        ggplot(aes(x=rank, y=distance,
                   label=gene_label)) +
                geom_point(size=2,
                           colour='steelblue') +
                geom_text_repel(force = 3) +
                theme_classic() +
                labs(x='Rank', y='Regulon distance (Gained Vs Reversed)')
dev.off()
```


```{r, creating_igraph_object}
stat1.gained <- filter(granet.interactions.df.list$gained, TF=='STAT1')
stat1.gained$category <- 'gained'
stat1.reversed <- filter(granet.interactions.df.list$reversed, TF=='STAT1')
stat1.reversed$category <- 'reversed'
stat1.reg <- rbind(stat1.gained, stat1.reversed)

nodes.df <- data.frame(gene=unique(c('STAT1', stat1.reg$gene)))
edges.df <- data.frame(from=stat1.reg$TF, to=stat1.reg$gene, category=stat1.reg$category)
g <- graph_from_data_frame(edges.df, directed = TRUE, vertices = nodes.df)
```


# STAT1 example


```{r, fig.height=10, fig.width=10}
selected.genes <- c('FOXP1', 'IRS2', 'PRDM1',
                    'TCF7', 'TNFRSF19', 'GZMA',
                    'HAVCR2')
labels <- sapply(V(g)$name, function(x){ ifelse(x %in% selected.genes, x, '') })

colors <- plyr::mapvalues(
        E(g)$category,
        from = unique(E(g)$category),
        to = c('salmon', 'green')
) 

l = layout_nicely(g)

pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/stat1_rewiring.pdf')
plot(g,
     vertex.size=1.5,
     vertex.label.dist=1,
     vertex.label.color='black',
     vertex.label.cex=0.5,
     
     edge.arrow.size=0.5,
     edge.curved=.1,
     edge.color=colors,
     
     layout=l)
dev.off()
```

# Ratio of regulon sizes

```{r}
interactions.list <- lapply(names(granet.interactions.df.list), 
                            function(name)
                            mutate(granet.interactions.df.list[name][[1]], category=name))
interactions.df <- do.call(rbind, interactions.list)

freq.df <- interactions.df %>%
        select(TF, category) %>%
        table() 
top.50 <- freq.df %>% 
        apply(1, sum) %>%
        sort(decreasing = TRUE) %>%
        head(50) %>%
        names()

ratios <- freq.df  %>%
        as.matrix() %>%
        apply(1, function(x) x/sum(x)) %>%
        as.data.frame() %>%
        t() %>%
        as.data.frame()

order.regs <- arrange(ratios, desc(scarred)) %>%
                rownames()


pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/ratio_interactions_by_category.pdf',
    width = 5, height = 7)
freq.df %>%
        as.data.frame() %>%
        filter(TF %in% top.50) %>%
        mutate(TF=factor(TF, levels = order.regs)) %>%
                ggplot(aes(x=TF, y=Freq, fill=category)) +
                        geom_bar(position = 'fill', stat = 'identity') +
                        coord_flip() +
                        theme_bw() +
                        theme(panel.grid = element_blank()) +
                        labs(x='Regulon', y='Ratio of interactions',
                             fill='')  
dev.off()

pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/freq_interactions_by_category.pdf',
    width = 5, height = 7)
freq.df %>%
        as.data.frame() %>%
        filter(TF %in% top.50) %>%
        mutate(TF=factor(TF, levels = rev(top.50))) %>%
                ggplot(aes(x=TF, y=Freq, fill=category)) +
                        geom_bar(position = 'stack', stat = 'identity') +
                        coord_flip() +
                        theme_bw() +
                        theme(panel.grid = element_blank()) +
                        labs(x='Regulon', y='Number of interactions',
                             fill='')  
dev.off()
```


# Number of conserved interactions

```{r}
interactions.df <- interactions.df %>%
        mutate(interaction=paste0(gene, '--',TF))
int.list <- split(interactions.df$interaction, interactions.df$category) 
ggVennDiagram(int.list)
```



# Visualisation of genes in selected regulons

## REL regulon


```{r, fig.width=8, fig.height=12}
path2seurat <- '/media/ag-cherrmann/cramirez/tcd8ExscSeq/data/maike2020/hofmann_hcv_seu.rds'
seurat <- read_rds(path2seurat)
seurat$nina_annotations %>% table()

DefaultAssay(seurat) <- 'RNA'

rel.reg <- filter(interactions.df, TF == 'REL' )  
rel.reg.list <- split(rel.reg$gene, rel.reg$category)
seurat$'condition' <- ifelse(is.na(seurat$nina_annotations), 'chronic', 'cured')
dotplot.hub0709 <- DotPlot(subset(seurat, orig.ident %in% c('HUBF0709BL', 'HUBF0709FUDay2')), 
        features = rel.reg.list$reversed, 
        group.by = 'orig.ident',
        cols = 'RdYlGn', 
        dot.scale = 10) +
        coord_flip() +
        ggtitle('HUBF0709') +
        labs(x='', y='') +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45,
                                         hjust = 1),
              panel.grid = element_blank())

dotplot.hubf004 <- DotPlot(subset(seurat, 
                                  orig.ident %in% c('HUBF0044BL', 'HUBF0044SORT3BFU')), 
        features = rel.reg.list$gained, 
        group.by = 'orig.ident',
        cols = 'RdYlGn', 
        dot.scale = 10) +
        coord_flip() +
        ggtitle('HUBF0044') +
        labs(x='', y='') + 
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45,
                                         hjust = 1),
              panel.grid = element_blank())


pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq//figures/gene_expression_REL_regulon.pdf',
    width = 8, height = 12)
dotplot.hub0709 + dotplot.hubf004
dev.off()



#####
rel.reg <- filter(interactions.df, TF == 'STAT1' )  
rel.reg.list <- split(rel.reg$gene, rel.reg$category)
seurat$'condition' <- ifelse(is.na(seurat$nina_annotations), 'chronic', 'cured')
dotplot.hub0709 <- DotPlot(subset(seurat, orig.ident %in% c('HUBF0709BL', 
                                                            'HUBF0709FUDay2')), 
        features = rel.reg.list$gained, 
        group.by = 'orig.ident',
        cols = 'RdYlGn', 
        dot.scale = 10) +
        coord_flip() +
        ggtitle('HUBF0709') +
        labs(x='', y='') +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45,
                                         hjust = 1),
              panel.grid = element_blank())

dotplot.hubf004 <- DotPlot(subset(seurat, 
                                  orig.ident %in% c('HUBF0044BL', 'HUBF0044SORT3BFU')), 
        features = rel.reg.list$gained, 
        group.by = 'orig.ident',
        cols = 'RdYlGn', 
        dot.scale = 10) +
        coord_flip() +
        ggtitle('HUBF0044') +
        labs(x='', y='') + 
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45,
                                         hjust = 1),
              panel.grid = element_blank())


pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq//figures/gene_expression_STAT1_regulon.pdf',
    width = 8, height = 15)
dotplot.hub0709 + dotplot.hubf004
dev.off()


```


```{r, hubf0709, fig.width=4.5, fig.height=15}
library(ComplexHeatmap)
seurat$orig.ident %>% table()

rel.reg.df <- filter(interactions.df, TF == 'REL')
rel.reg.df
targets <- split(rel.reg.df$gene, rel.reg.df$category) %>% unlist()
names(targets) <- gsub('[[:digit:]]*$', '', names(targets))
mtx.hubf004 <- FetchData(subset(seurat, orig.ident %in% c('HUBF0709BL', 
                                                       'HUBF0709FUDay2')), 
                         vars = rel.reg.df$gene)
mtx.hubf004 <- t(mtx.hubf004) %>% as.matrix()

gene.anns <- rel.reg.df

patient.anns <- FetchData(subset(seurat, orig.ident %in% c('HUBF0709BL', 
                                                       'HUBF0709FUDay2')), 
                         vars = 'orig.ident')

## calculating top HVG 
top40hvgGenes <- mtx.hubf004 %>%
        apply(1, sd) %>%
        sort(decreasing = TRUE) %>%
        head(50) %>%
        names()

#gene.anns <- filter(gene.anns, gene %in% unique(top40hvgGenes))

## only non-duplicated genes
rel.reg.df <- filter(interactions.df, TF == 'REL') %>%
                filter(!duplicated(gene))

mtx.hubf004 <- FetchData(subset(seurat, orig.ident %in% c('HUBF0709BL', 
                                                       'HUBF0709FUDay2')), 
                         vars = c(rel.reg.df$gene, 'orig.ident')
                         )
mtx.hubf004 <- mtx.hubf004 %>%
        group_by(orig.ident) %>%
        summarise_all(mean) %>%
        column_to_rownames('orig.ident') %>%
        t() %>%
        as.matrix()


mtx.hubf004 <- mtx.hubf004[rel.reg.df$gene, ]

pdf('/media/ag-cherrmann/cramirez/tcd8ExscSeq/figures/rel_regulon_by_category_hubf0709.pdf',
    height = 18, width = 5)
Heatmap(
        mtx.hubf004, 
        show_column_names = TRUE, 
        row_split = rel.reg.df$category, 
        col = viridis(20)
)
dev.off()
```


```{r, hubf0709, fig.width=4, fig.height=15}
library(ComplexHeatmap)
seurat$orig.ident %>% table()

rel.reg.df <- filter(interactions.df, TF == 'BACH2')
rel.reg.df
targets <- split(rel.reg.df$gene, rel.reg.df$category) %>% unlist()
names(targets) <- gsub('[[:digit:]]*$', '', names(targets))
mtx.hubf004 <- FetchData(subset(seurat, orig.ident %in% c('HUBF0044BL', 
                                                       'HUBF0044SORT3BFU')), 
                         vars = rel.reg.df$gene)
mtx.hubf004 <- t(mtx.hubf004) %>% as.matrix()

gene.anns <- rel.reg.df

patient.anns <- FetchData(subset(seurat, orig.ident %in% c('HUBF0044BL', 
                                                       'HUBF0044SORT3BFU')), 
                         vars = 'orig.ident')

## calculating top HVG 
top40hvgGenes <- mtx.hubf004 %>%
        apply(1, sd) %>%
        sort(decreasing = TRUE) %>%
        head(50) %>%
        names()

#gene.anns <- filter(gene.anns, gene %in% unique(top40hvgGenes))

## only non-duplicated genes
rel.reg.df <- filter(interactions.df, TF == 'BACH2') %>%
                filter(!duplicated(gene))
rel.reg.df <- rel.reg.df[!duplicated(rel.reg.df$gene), ]

mtx.hubf004 <- FetchData(subset(seurat, orig.ident %in% c('HUBF0044BL', 
                                                       'HUBF0044SORT3BFU')), 
                         vars = c(rel.reg.df$gene, 'orig.ident')
                         )
mtx.hubf004 <- mtx.hubf004 %>%
        group_by(orig.ident) %>%
        summarise_all(mean) %>%
        column_to_rownames('orig.ident') %>%
        t() %>%
        as.matrix()


rel.reg.df <- column_to_rownames(rel.reg.df, var = 'gene')
intersect.genes <- intersect(rownames(rel.reg.df), rownames(mtx.hubf004))
mtx.hubf004 <- mtx.hubf004[intersect.genes, ]

Heatmap(
        mtx.hubf004, 
        show_column_names = TRUE, 
        row_split = rel.reg.df[intersect.genes,]$category, 
        col = viridis(20),
        column_dend_reorder = NA
)
```